%% The following is a directive for TeXShop to indicate the main file
%%!TEX root = ../thesis.tex

\chapter{Acknowledgments}

I am very fortunate to have been supported by a tremendous group of people during my academic and personal journeys as I conducted this research.

First and foremost, I am deeply grateful to my supervisor Dr. Doug Oldenburg -- your endless enthusiasm and desire to make a positive impact, both in the field of geophysics and the daily lives of those around you, have enriched my life. You have opened doors for me and encouraged me to pursue projects I find meaningful. For these things and more, thank you.

Next, I thank Dr. Eldad Haber; the numerical work in this thesis has its foundation in your courses and resources on computational electromagnetics. Thank you for all you have taught me. Thanks to Michael Bostock for serving on my committee and for your input which helped improve this work.

To Dr. Michael Wilt, conversations with you helped shape the questions I pursued and the final content of this thesis. I am grateful for the investment you have made in me.

There are numerous educators who have inspired me and motivated my interest in pursuing science. In particular, Linda Ledgenre, David Westra, and Dr. David Lawrie, you have each had a meaningful impact on my career.

A special thanks to the friends who have supported me through my PhD. To name a few: Dom Fournier, Sarah Devriese, Thibaut Astic, Gudni Rosenkjaer, Mike McMillan, Mike Mitchell, Devin Cowan, Luz Angelica Caudillo-Mata, Dikun Yang, Kris Davis, Dave Marchant, Jenn Fohring, Justin Granek, and Julie Nutini. I especially thank Rowan Cockett and Seogi Kang, I have learned so much from you both; I am grateful to count you as partners in much of this work and as supportive friends. To the growing number of contributors to SimPEG and resources in the open source ecosystem, I am inspired by your efforts.

To my family, who have been unwavering in their support including my encouraging brother, Josh Heagy, and my grandparents.

Finally, to my parents, I could not have done this without you. I am forever grateful for your love and support.

Funding for this work was provided by Vanier Canada Graduate Scholarship Program, the University of British Columbia Four Year Fellowship.
