%% The following is a directive for TeXShop to indicate the main file
%%!TEX root = ../../thesis.tex

\chapter{A physical property model for a fractured volume of rock}
\label{ch:phys_prop_model}

For electromagnetic methods to be sensitive to a propped, fractured volume of rock, the fractured volume of rock must have physical properties which are distinct from the background, host rock. For EM methods, this means the electrical conductivity, magnetic permeability, or dielectric permittivity of the fractured rock must be distinct. Dielectric permittivity only plays a significant role when the frequency of the source is sufficiently high, in the hundreds of kilohertz to megahertz range, as is used in ground penetrating radar. Over the length-scales we need to consider for imaging fractures, 100’s to 1000’s of meters, attenuation of the EM signals due to skin depth effects make GPR impractical. Thus, we will work at lower frequencies, in the quasi-static regime of Maxwell's equations, and concern ourselves only with magnetic permeability and electrical conductivity. The materials traditionally used as proppant, typically sand and ceramics, tend to have similar physical properties to the reservoir that they are being pumped into, making it difficult to detect them on the scale of the reservoir. However, if the proppant were made electrically conductive or magnetically permeable, for instance by coating it with graphite or including magnetite particles, it may create a sufficient physical property contrast that can be imaged using EM.

\subsection{Proppant selection}
To make the proppant electromagnetically distinct from the host rock, either the magnetic permeability or the electrical conductivity of the proppant could be targeted. \cite{Zawadzki2016} provides an overview of possible magnetic proppants and highlights some practical considerations for the choice of material. For any proppant or additive, the constraints include: the mechanical strength of the particles must be able to withstand the pressure of the closing fracture without crushing and clogging the fracture pathways; the material shouldn’t be toxic or reactive; and the price-point should be reasonable since high volumes are needed. \cite{Zawadzki2016} also provide a general classification of material types that could be considered: feedstock material, materials that are mixed with conventional proppant or replace conventional proppant, could include magnetite or steel particles. Both have a significant magnetic permeability, however, magnetite crushes easily, and although steel has significant mechanical strength, it is challenging to manufacture particles that are small enough (typically $< 2$mm in diameter). They also consider ferrofluids, which contain microscopic ferromagnetic particles in suspension, and magnetic nanoparticles. Both are sufficiently small and remain in suspension so clogging is less of a concern. However, the cost of either material is quite significant and steps must be taken to reduce the environmental hazard posed. This is particularly important for nanoparticles as they tend to be much more reactive than particles of the same material but larger in size \citep{Zawadzki2016}. Continuing advances in nanotechnology has prompted some authors, e.g. \citep{Rahmani2014}, to pursue analysis of the use of magnetic markers for mapping hydraulic fractures using electromagnetic techniques.

Magnetic permeability of common materials tends to vary over an order of magnitude \cite{Telford1990}, while electrical conductivity of common materials can vary over $> 8$ orders of magnitude. Comparatively, there are many more materials that are electrically conductive than there are with a significant magnetic permeability. Materials such as coke breeze, a hardened graphite coating applied to conventional proppants, have been considered by numerous authors, as have a range of manufactured electrically conductive proppants \citep{Pardo2013, Hoversten2015, Weiss2015, Labrecque2016, Hu2018}. For these reasons, we focus on electrically conductive proppants and treat the fractured region of the reservoir as an electrically conductive geophysical target.

\subsection{Numerical Modelling}
Numerical modelling is a critical component for assessing the feasibility of detecting a fractured volume of rock with an electromagnetic survey. To run a simulation, we need to discretize the modelling domain and represent the electrical conductivity of the earth on the simulation mesh. It is important to consider the large range of scales at play when considering fractures. The proppant which fills the fractures is micro-to-millimeters in diameter. The fractures are millimeters thick and we are aiming to characterize a region of the reservoir which extends hundreds of meters from the injection point at the well, tens to hundreds of meters in height, and tens of meters along the length of the well-bore. Furthermore, the numerical simulation domain must extend far enough to satisfy boundary conditions (typically that the fields have sufficiently decayed). This problem is exacerbated when one considers an inversion, where many forward simulations must be performed. We cannot expect a method to be capable of imaging both such a substantial volume of the subsurface while also having the resolution on the scale of the proppant particles. Thus, we require a characterization of the bulk impact of the conductive proppant within a fractured volume of rock. How we construct such a bulk physical property model depends on the geometry and complexity of the induced fractures, for instance, different approaches should be considered if the fracture is a simple planar fracture versus a complex fracture network, as depicted in Figure \ref{fig:frac_complexity} (c.f. \cite{Cipolla2008}).

To overcome these challenges, there are several approaches that may be taken; the appropriate choice will depend upon both the fracture complexity and the purpose of the simulation. If the fine-scale geometry of the fractures is defined, for example in a feasibility study built from a synthetic fracture model, then numerical upscaling \citep{Durlofsky2003, CaudilloMata2014, CaudilloMata2016} or multiscale techniques \citep{Haber2014} can be employed. In general, though, the fine-scale fracture geometry is not known; only a handful of studies have ``ground-truthed'' the geometry of an induced fracture by mining the fractured region \citep{Cipolla2008a}. Furthermore, in an inversion, we cannot expect to resolve individual fractures, rather, we aim to characterize the bulk impact due to electrically conductive fractures in a conductive medium. To estimate a bulk conductivity, effective medium theory is one approach that can be taken (c.f. \cite{Torquato2002, Milton2002, Berryman2013}). These methods assume that the composite material is composed of a collection of randomly distributed spheres or ellipsoids (which may be preferentially aligned). The estimated conductivity depends upon the conductivity of each of the materials, their shape, and volume in the composite material. Looking forward to the inverse problem, these methods have the added benefit that they provide a conduit for incorporating a-priori information such as expected primary-fracture orientation, and volumes of proppant and fluid.

In this work, we adopt an effective medium theory approach. We take into account the conductivity of the host rock, the fracturing fluid and the proppant. The main simplifying assumption we employ is on the geometry of the fractures: we assume that they are composed of a collection of ellipsoidal cracks which may be preferentially or randomly aligned.
\subsection{Chapter overview}
The purpose of this chapter is twofold: (1) we develop a workflow for estimating the physical properties of a fractured volume of rock containing proppant and fluid that have distinct electromagnetic properties from the host and (2) we assess if, using a commercially available electromagnetic survey, we can detect the anomaly introduced by the fracture in the data. All of the computations shown in this chapter are open-source and available as Jupyter notebooks at: https://github.com/simpeg-research/heagy-2018-fracture-physprops.

\section{Homogenization workflow using effective medium theory}

In this approach, we treat the fractures as being composed of a collection of preferentially (or randomly) oriented ellipsoidal cracks and based on the density of cracks within a given volume of rock, construct an anisotropic description of the coarse-scale conductivity.

Effective medium approximations range from applying simple harmonic or arithmetic averaging of conductivity values, for example when constructing a representative voxel conductivity model from well-log measurements, to more involved analytic or empirical relationships such as Archie’s law \citep{Archie1942}, which is commonly applied for estimating the conductivity of a fluid-filled rock. Although discovered empirically, the simplest version of Archie’s law is one example of a differential effective medium approximation which can be derived analytically. It assumes a background matrix, and uses an incremental approach to constructing a homogenized conductivity (c.f. \cite{Torquato2002, Milton2002}). However, for describing a fractured volume of rock, differential effective medium approximations are not appropriate as they assume that the rock-matrix is always connected \citep{Torquato2002}; in the composite material we are considering, a single computational voxel may be intersected by a fracture, meaning the rock matrix is not connected. The Maxwell-approximation \citep{Maxwell1873} is yet another effective-medium approximation. It again makes a distinction between the background and the included phases, and assumes no interaction between inclusions.

We opt to consider self-consistent effective medium theory (SCEMT, also sometimes referred to as the Coherent Potential Approximation, CPS, or Bruggeman mixing). This is an effective medium approach which makes no distinction between background and included phases (e.g. see \cite{Torquato2002}). \cite{Berryman2013} similarly suggest using self-consistent effective medium theory for fractured rocks where the fractures are filled with fluid.

\cite{Milton1985} demonstrated the physical realizability of SCEMT; he showed that SCEMT is asymptotically the exact solution for the effective conductivity for fractal-like composites that are self-similar at many scales. Physical realizability means that the estimates given by SCEMT will always be within the Hashin-Shtrikman bounds; at a minimum, this provides confidence that the conductivity estimates it provides are physically within a reasonable relm.

We construct the physical property model for a fractured volume of rock in two steps, as shown in Figure \ref{fig:effective_medium_theory}. Given the conductivity of the fluid and proppant particles, we estimate the effective conductivity, $\sigma_2$, of a mixture of proppant and fluid. Next, we treat the fracture as consisting of a collection of ellipsoidal cracks filled with the proppant-fluid mixture. The ellipsoidal cracks may be preferentially aligned in a single or multiple directions or they may be randomly oriented. In both cases, we use the self-consistent effective medium theory, originally due to \cite{Bruggeman1935} and further developed by \cite{Landauer1952, Landauer1978}. In the following sections, we develop the theory and demonstrate its application for computing the effective conductivity of a proppant-fluid mixture as well as a fractured volume of rock.


\input{figures/phys_prop_model/effective_medium_theory}



\subsection{Summary of self-consistent effective medium theory}
\label{sect:emt_math}

Chapter 18 of \cite{Torquato2002} provides an overview of effective medium theory approaches, and the discussion presented in this section follows their presentation. Here, we discuss self-consistent effective medium theory, originally developed by \cite{Bruggeman1935}. \cite{Berryman2013} present a similar overview, introducing several simplifying assumptions tailored for estimating the effective conductivity of a naturally fractured rock where the fractures are a relatively small concentration with respect to the host rock. Here, we work with the general formulation with the aim of being suitable for calculating both the effective conductivity of a proppant-fluid mixture as well as arbitrarily oriented fractures, each at potentially high concentrations.

Each material, or phase, in the composite is assumed to be made up of spherical or ellipsoidal particles with a known aspect ratio. Starting from the solution for a sphere or an ellipsoid in a uniform electric field, the effective conductivity of a heterogeneous medium is chosen to be the conductivity for which the average perturbation to the electric field -- the difference between the electric field in the homogenized medium and the true conductivity model -- is zero. That is,
\input{equations/effective_medium_theory}
where $N$ is the number of different phases of materials, $\varphi_j$ is the volume fraction of the $j$-th phase, and $\sigma_j$ is the electrical conductivity of the $j$-th phase. $\boldsymbol{\Sigma^*}$ is the $3 \times 3$ effective conductivity tensor, and $\mathbf{I}$ is the $3 \times 3$ identity matrix. The matrix $\mathbf{R}^{(j,*)}$ is the electric field concentration tensor, and depends both on the shape of the inclusions (ie. proppant particles or cracks composing a fracture) and conductivity of the $j$-th phase, as well as the effective conductivity $\boldsymbol{\Sigma^*}$.

For spherical particles, the electric field concentration tensor $\mathbf{R}^{(j,*)}$ reduces to a scalar, namely,
\input{equations/emt_r_spheres}
The resultant effective conductivity expression in \ref{eq:effective_medium_theory} then reduces to a scalar equation.

If instead ellipsoidal inclusions are considered, the electric field concentration tensor is given by
\input{equations/emt_r_ellipsoids}
Where $\mathbf{A}$ is the de-polarization tensor. For simplicity, we will assume that we are working with spheroids, either oblate (pancake-like) or prolate (needle-like) spheroids which have two semi-axes that are equal. The general solution for spheroids with three distinct semi-axes is presented in chapter 17 of \cite{Torquato2002} and additionally discussed in \cite{Berryman2013}. For a spheroid with semi-axes $a_1 = a_2 = a$ and $a_3 = b$ that is aligned so that $a_3$ lies along the z-axis, the depolarization tensor is given by
\input{equations/emt_depolarization_tensor}
For prolate spheroids, with aspect ratio $\alpha = b/a > 1$, $Q$ is given by
\input{equations/emt_q_prolate_spheroid}
and for oblate spheroids
\input{equations/emt_q_oblate_spheroid}
with
\input{equations/emt_chi}
For preferentially aligned spheroids, the matrix $\mathbf{A}$ can be rotated as to align with the using standard coordinate rotations. If the spheroids are randomly oriented, then we replace $\mathbf{R}^{(j,*)}$ with $1/3\text{trace}(\mathbf{R}^{(j,*)})$ in equation \ref{eq:effective_medium_theory}, and the effective conductivity expression reduces to a scalar equation. Note, that for spherical inclusions, $Q = 1/3$ and thus $\mathbf{A}$ reduces to a scalar equal to 1/3, showing that \ref{eq:emt_r_ellipsoids} is consistent with \ref{eq:emt_r_spheres}.
To solve for the effective conductivity, which is an implicit expression for $\Sigma^*$, we rearrange equation \ref{eq:effective_medium_theory} to
\input{equations/emt_solving_a}
and solve
\input{equations/emt_solving_b}
using fixed-point iteration as $\mathbf{R}^{(j, *)}$ depends on $\Sigma^*$. Note that the matrix inverse in \ref{eq:emt_solving_b} is a $3\times 3$ matrix inverse and thus is cheap to explicitly compute. The fixed point iteration is performed until the recovered effective conductivity converges within a predefined tolerance.\footnote{\cite{Berryman2013} similarly use a fixed-point iteration to solve for the effective conductivity, however in their formulation, they isolate $\Sigma^*$ by pulling out the first term in the summation in \ref{eq:effective_medium_theory}, leading to an update of the form
\input{equations/emt_berryman_update.tex}
This is equivalent to equation 10 in \cite{Berryman2013} with the simplifications that $\mathbf{R}^{(j, *)}$ is replaced by $\mathbf{R}^{(j, 0)}$ under the assumption that the fractures compose a small volume fraction of the fractured rock. In practice, this approach is suitable for low concentrations of included phases, but can cause instability in the algorithm at higher concentrations (it is possible for updates to be negative, and therefore unphysical); the authors noted challenges with algorithm convergence when the concentration of inclusions exceeded 0.2.} In the following sections, we use this formulation to estimate the effective conductivity of a range of proppant-fluid mixtures as well as for a fractured volume of rock.

\subsection{Step 1: Effective conductivity of a proppant-fluid mixture}
In general, an induced fracture will be filled with two types or phases of material: proppant and fluid. Often this mixture is referred to as a slurry. There are two principal types of mixtures we will consider, one where all of the proppant is conductive, and the second when there is a conductive filler added to a conventional proppant.

We start by considering the case of a proppant with uniform conductivity, for instance if a conductive proppant, or coated proppant were used. We assume the conductivity of the proppant particles is known; Appendix \ref{app:concentric_spheres} includes a derivation for an effective conductivity of two concentric spheres for the scenario where the conductivity of a proppant particle and its coating are known independently.

For spherical proppant particles, the tensor-values in equation \ref{eq:effective_medium_theory} reduces to a scalar-valued equation, and the resulting effective conductivity is isotropic. In Figure \ref{fig:emt_spherical_particles}, we show the effective conductivity found using equations \ref{eq:effective_medium_theory} and \ref{eq:emt_r_spheres} for a proppant-fluid mixture as the concentration of proppant in the mixture ($\varphi$) is varied. The conductivity of the fluid is 3 S/m (similar to that of sea-water), and the proppant conductivity is varied logarithmically from 10 S/m to $10^5$ S/m. For example, coke-breeze, a graphite based material has conductivities $\sim 3000$ S/m, and other authors have considered the use of contrast agents that reach conductivities of $10^5$ S/m \cite{Weiss2015} and $10^6$ S/m \cite{Pardo2013}, which are similar to the conductivity of steel.


\input{figures/phys_prop_model/emt_spherical_particles}

When the volume fraction of proppant is less than $1/3$, the conductivity of the fluid dominates is the dominant control on the resulting effective conductivity. Above a volume  fraction of $1/3$, the conductivity of the proppant is the primary contributor to the effective conductivity. The threshold between these behaviors is the \emph{percolation threshold}. Below it, the concentration of conductive material is low enough that it is quite likely disconnected, above $1/3$, the concentration is high enough to start forming connected, electrically conductive pathways, causing a large jump in the effective conductivity of the system. Although proppant typically composes $10\%$ to $20\%$ of the injected slurry, some of the injected fluid leaks off into the surrounding geologic formation leaving proppant concentration that can be $50\%$ in the fractures \cite{Novotny1977, Hoversten2015}.

The conductivity of the fluid also changes the resultant effective conductivity of the mixture. In Figure \ref{fig:emt_fluid}, we compare the effective conductivity for a mixture of conductive proppant (panel (a): $10^3$ S/m, panel (b): $10^4$ S/m) and four different fluid conductivities, ranging from 0.3 S/m to 300 S/m, as indicated in the legend. Although the conductivity of the fluid makes a significant difference at low proppant concentrations, above the percolation threshold of 33\%, the curves start to converge, particularly when the contrast between the conductivity of the proppant and the fluid exceeds 3 orders of magnitude. Thus, if the proppant can be made sufficiently conductive, its conductivity will be the controlling factor on the effective conductivity of the slurry that remains in the fractures.


\input{figures/phys_prop_model/emt_fluid}

The previous examples considered a 2-phase mixture in which all of the proppant was electrically conductive, however, depending on the setting and the cost to manufacture conductive proppant, it may be mixed in with conventional, resistive proppant. To examine this, we consider a proppant-fluid mixture composed of three materials, fluid ($3$ S/m), conventional, resistive proppant ($10^{-6}$ S/m) and conductive proppant ($10^5$ S/m). The effective conductivities of proppant-fluid mixtures for five different proppant blends, where the relative concentration of the conductive proppant is varied from 0\% to 100\% of the proppant phase, are shown in Figure \ref{fig:emt_3phase}. Again, we see the impacts of the percolation threshold; when the conductive proppant composes less than $1/3$ of the proppant pack, the effective conductivity is dominated by the resistive proppant. When the conductive proppant composes more than $1/3$ of the proppant pack, we see that with increasing proppant concentration, the effective conductivity of the mixture is dominated by the conductive proppant. However, the percolation threshold for each of these mixtures is different. This is because it is the role of the conductive proppant in the three-phase mixture, not the ratio of proppant to fluid, that determines when connected, electrically conductive pathways may be formed.


\input{figures/phys_prop_model/emt_3phase}

Another factor influencing the effective conductivity of a mixture is the shape of the materials. For the previous examples, the proppant was assumed to be composed of spherical particles. If elongated, conductive particles were included, we expect that connected, conductive pathways would form at lower concentrations. For instance, consider a 3 phase proppant mixture consisting of fluid (3 S/m), resistive, spherical proppant ($10^{-6}$ S/m), and elongated, electrically conductive proppant ($10^5$ S/m). Assume that the ratio of conductive to resistive proppant is 0.25 (below the percolation threshold for spherical particles). If the elongated particles (prolate spheroids) are randomly oriented, then the resulting effective conductivity is isotropic, meaning it is independent of the directions of the inducing field and resulting current. The conductivity predicted by effective medium theory for mixtures with five different aspect ratios is shown in figure \ref{fig:emt_3phase_aspect}. The aspect ratio of the conductive particles influences the concentration at which we observe percolation. The more elongated the particles, the lower the concentration at which percolation occurs.



\input{figures/phys_prop_model/emt_3phase_aspect}

In summary, there are several approaches that can be taken to create an electrically conductive proppant-fluid mixture. Spherical proppant particles which are themselves electrically conductive or coated with a conductive material can comprise the entire proppant pack. If conductive proppant is mixed with a conventional, resistive sand or ceramic particles, then at least 1/3 of the proppant mixture needs to be comprised of electrically conductive particles to create a conductive mixture. This ratio can be reduced if elongated particles, such as metallic strips, are used in the proppant mixture.

\subsection{Step 2: Effective conductivity of fractured volume of rock}
The next step is to estimate the effective conductivity of a fractured volume of rock. We again employ self-consistent effective medium theory as described in section \ref{sect:emt_math} and consider the induced fractures to be composed of spheroidal cracks. Based on the analysis in the previous section, we consider a proppant-fluid mixture that has a conductivity of 2500 S/m. This could be achieved with spherical proppant particles having a conductivity of $10^4$ S/m in a 50/50 mixture with water of 3 S/m (see Figure \ref{fig:emt_spherical_particles}). Similar conductivities could be achieved with elongated particles mixed with conventional proppant as shown in Figure \ref{fig:emt_3phase_aspect}. Lab measurements conducted by \cite{Zhang2016} showed that a proppant-fluid mixture composed of petroleum coke particles and seawater which fills the pore-spaces reached a conductivity of $\sim 1000$ S/m at 37.6\% porosity. With further increase in confining pressure (thus reducing porosity and increasing the concentration of proppant), the observed conductivities of $3000 - 5000$ S/m. These results provide further confidence that conductivities $> 1000$ S/m for the mixture filling the hydraulic fractures are attainable.


For the following example, we will assume that the conductivity of the host-rock is 0.1 S/m. There are two scenarios we will consider, in the first, we assume the cracks are preferentially aligned, with the thin dimension of the oblate spheroid oriented along the y-axis. In this case, the recovered effective conductivity will be anisotropic, described by a diagonal matrix with entries $\sigma_x = \sigma_z \geq \sigma_y:$
\input{equations/diagonal_sigma_e}
Note that arbitrary, non-axes aligned, orientations can be considered; all that is required is that the depolarization tensor described in \ref{eq:emt_depolarization_tensor} is rotated to the desired orientation.

To estimate the effective conductivity of a fractured volume of rock, we must also specify the aspect ratio of the cracks. In estimating this, assume a fractal-like approximation, where the aspect ratio of the fracture is representative of the aspect ratio of the cracks that compose it. For example, if the fracture extends 50m laterally and has a width on the order of millimeters, then the aspect ratio is on the order of $10^{-5}$. In Figure \ref{fig:aligned_fractures}, we have plotted the diagonal elements of the effective conductivity for five different aspect ratios, indicated in the legend, as a function of the volume fraction of conductive fractures in the rock volume sampled. Panel (a) shows the full range from $0 \leq \varphi \leq 1$ and panel (b) zooms in to lower concentrations ($0 \leq \varphi \leq 0.01$) which are more representative of a fractured rock volume, on the scale that we will consider for numerical modelling (e.g. if 10 fractures, each with 3mm width intersect a 10m $\times$ 10m $\times$ 10m cell, then $\varphi = 0.003$). In each of the plots, we have also included the upper and lower Wiener bounds (see equation 21.14 in \cite{Torquato2002}; originally attributed to \cite{Wiener1912}):
\input{equations/wiener_bounds}
in the black dashed lines; these can be understood as similar to parallel and series circuit approximations to the conductivity. In the black dash-dot lines are the upper and lower Hashin-Shtrikman bounds for 2-phase anisotropic media in the black dash-dot (see equations 21.25 and 21.26 in \cite{Torquato2002}, which is the anisotropic generalization of the isotropic bounds derived by \cite{Hashin1962}):
\input{equations/hashin_shtrikman_bounds_anisotropic}
For $\sigma_1 \geq \sigma_0$ and
\input{equations/hashin_shtrikman_bounds_A}
where $\mathbf{A}_1$ is the depolarization tensor for the ellipsoidal cracks given by equation \ref{eq:emt_depolarization_tensor}. For the bounds shown in the plot, the smallest aspect ratio, $10^{-5}$ was used to calculate the depolarization tensor. For the very significant aspect ratios used here, the Hashin-Shtrikman bounds are nearly identical to the Wiener bounds. Each component of the recovered effective conductivity should fall within these bounds.

\input{figures/phys_prop_model/aligned_fractures}

For aspect ratios less than $10^{-3}$, we see very little distinction between the estimate of $\sigma_{x, z}$ and $\sigma_{y}$; the difference between the recovered effective conductivities for the aspect ratios of $10^{-4}$ and $10^{-5}$ is less than 1\% for all values of $\varphi$. This indicates that for sufficiently thin cracks, the exact aspect ratio is not critical for estimating a representative conductivity of the fractured rock.

At the concentrations we expect to be observing in a hydraulic fracturing scenario (Figure \ref{fig:aligned_fractures}b), we see that the effective conductivity along the normal of the fractures coincides with the lower bounds and remains nearly identical to the conductivity of the host rock for all aspect ratios shown, as may be expected. For the components of the conductivity along the cracks ($\sigma_{x, z}$), the effective conductivity mimics the behavior of the upper bounds. These components of the conductivity are similar to the behavior expected from a parallel-circuit approximation.

For settings where planar fractures are expected, an single orientation of the inclusions produces similar behavior to what would be expected if we performed simple series-and-parallel circuit approximations to the components perpendicular to and along the fracture. However, if more complex fracture networks are expected, it may be more appropriate to assume the cracks are randomly oriented. In this case, an isotropic conductivity describes the fractured volume of rock. Using the same aspect ratios shown in Figure \ref{fig:aligned_fractures} above, we compute the isotropic effective conductivity as a function volume fraction of fractures in Figure \ref{fig:random_fractures}. For reference, the conductivity along the fractures, $\sigma_{x, z}$ from Figure \ref{fig:aligned_fractures}, is plotted in semi-transparent lines. As in the previous Figure, panel (a) shows the full range from $0 \leq \varphi \leq 1$ and panel (b) zooms in to lower concentrations: $0 \leq \varphi \leq 0.01$.

\input{figures/phys_prop_model/random_fractures}

Again, we see that for aspect ratios smaller than $10^{-3}$ there is little difference between the estimated effective conductivity of the fractured rock volume. The estimate of the effective conductivity largely follows the behavior of the upper bounds, while the magnitude of the effective conductivity is slightly reduced as compared to the component of the conductivity along the fractures in the anisotropic case. If we consider a $10m \times 10m \times 10m$ computational cell with fractures having an aspect ratio of $10^{-5}$ and composing 0.3\% of the total volume (e.g. 10 fractures, each with 3mm width), we obtain $\sigma_x = \sigma_z = 5.2$ S/m, $\sigma_y = 0.1$ S/m for the case where the cracks are preferentially aligned while for the case where the cracks are randomly oriented, we obtain $\sigma^* = 3.5$ S/m.

Due to the large contrast between the conductive fractures and the host rock, the conductivity of the background has marginal effect on $\sigma_{x, z}$ in the anisotropic example or on $\sigma^*$. If we consider the example with $\varphi = 0.003$, as before, and instead use a background conductivity of 0.01 S/m, the effective anisotropic conductivity is $\sigma_x = \sigma_z = 5.1$ S/m, $\sigma_y = 0.01$ S/m and for the isotropic case $\sigma^* = 3.4$ S/m. However, if the background is made more conductive and the contrast between the conductive fractures and the host reduced, we do see some impact. Setting the background to 1 S/m for this same example, we obtain $\sigma_x = \sigma_z = 6.4$ S/m, $\sigma_y = 1$ S/m for the anisotropic case and $\sigma^* = 4.7$ S/m for the isotropic case.


\subsection{Summary}
To construct an approximate conductivity model of a fractured volume of rock, we use a two step process: first, we estimate the effective conductivity of a mixture of proppant and fluid that fills the fractures and second, we estimate the effective conductivity of a fractured volume of rock.

In the section that follows, we will compare numerical simulations of a crosswell electromagnetic survey for both anisotropic and isotropic conductivity models of a fractured volume of rock, representing planar and complex fracture networks, respectively.

\section{Feasibility of EM for detecting fractures}
With the capability to estimate a physical property model of a fractured volume of rock, will now examine the feasibility of detecting a conductive, fractured region of a reservoir with an electromagnetic survey. The purpose of this section is to demonstrate detectability for a simple model of a fracture with existing electromagnetic imaging techniques. The survey we consider is a frequency-domain crosswell electromagnetic survey. In a crosswell EM survey a transmitter coil, which produces a time-varying magnetic field at a single frequency is positioned inside of one well. Time varying magnetic fields induce time-varying currents in the earth whose distribution and magnitude depends upon the conductivity of the earth. These currents in turn generate secondary magnetic fields. In a second well, an array of receiver induction coils, which measure the magnetic field, are positioned. To conduct a survey, the position of the receivers is fixed and the transmitter is moved at a fixed rate along the borehole (typically 3-5 meters per minute). The receivers are repositioned, and the process repeated \citep{Wilt1995}. Both the transmitter and receivers are oriented along the axis of the borehole; for vertical wells, this means the transmitter is a vertical magnetic dipole, and the receivers measure the z-component of the magnetic field. Since its development, crosswell surveys have been used for a range of reservoir imaging problems for enhanced oil recovery including monitoring steam injections \citep{Wilt1997} and water floods \citep{Wilt2012}, including a survey conducted in a pair of horizontal wells \citep{Marsala2015, Marsala2015a}. In this section, we consider a crosswell survey and examine the feasibility of detecting signal due to a fractured volume of rock.

The setup we consider is shown in Figure \ref{fig:crosswell_fractures}; a crosswell EM survey conducted in a pair of horizontal wells that are spaced 500m apart, as shown in Figure \ref{fig:crosswell_fractures}. The transmitter is oriented along the x-axis and positioned the same well as where the fracture stimulation was performed; the receivers are in the offset well and measure the real and imaginary parts of the x-component of the magnetic field. In the numerical modelling, we assume that the wells are sufficiently deep so that the background can be treated as a whole-space. We do not consider steel-cased boreholes in this example and instead assume that the boreholes are open-holes or cased with fibreglass casing as in \cite{Wilt2012}. The additional complications due to steel cased wells will be discussed in the following two chapters.

Most hydraulic fracture operations involve injecting $\sim$800m$^3$ to 4000 m$^3$ (5000 to 25000 bbls) of proppant-fluid slurry into the reservoir rock, with the proppant comprising 10\% to 20\% of this mixture \citep{Hoversten2015}. In this example, we consider a fracture that is on the lower-end, with an 800m$^3$ slurry comprised of $15\%$ proppant by volume. A significant portion of the fluid typically leaks off into the surrounding formation, so we will assume that the final mixture filling the cracks contains 50\% fluid and 50\% proppant. This gives a total fracture volume of 240 m$^3$. To distribute this volume, we assume that 10 fractures, each with 3mm width are positioned within a 10m interval along the well. Conserving volume and assuming circular fractures gives us a 50m radius for each of the 10 fractures; we work with circular fractures as the resultant model is cylindrically symmetric, which significantly reduces the computational cost of the simulation.

\input{figures/phys_prop_model/crosswell_fractures}

For the physical properties, we treat the background as a whole-space with a conductivity of 0.1 S/m. Within the fractures, we use a two-phase proppant-fluid mixture, where all of the proppant are assumed to be spherical particles as in Figure \ref{fig:emt_spherical_particles}. We use a proppant conductivity of $10^4$ S/m and a fluid conductivity of 3 S/m; in a 50-50 mixture, the effective conductivity computed using SCEMT, is 2500 S/m. To estimate the effective conductivity of the fractured volume of rock, we consider two scenarios: (1) the fractures are preferentially aligned and we estimate an anisotropic effective conductivity as in Figure \ref{fig:aligned_fractures} and (2) the fractures are sufficiently complex that we can approximate them as a collection of randomly oriented cracks, as in Figure \ref{fig:random_fractures}. We estimate a single conductivity for the 10m wide cylinder with 50m radius that encloses the fractures. Within this region, the fractures comprise 0.3\% of the total volume. To estimate the aspect ratio, we assume that the shape of the large, with 3mm width and 50m radius (100m diameter) is representative of the shape of the cracks that comprise it, thus the aspect ratio is $3\times10^{-5}$. The estimated effective conductivity for the anisotropic case is $\sigma_y = \sigma_z = 5$ S/m parallel to the fractures and $\sigma_x = 0.1$ S/m perpendicular to the fractures. For the second scenario, with randomly oriented cracks, the estimated isotropic conductivity is 3 S/m.

In terms of survey parameters, the dipole-moment of the transmitter is 5000 Am$^2$ and we use a transmitter frequency of 100Hz, consistent with values stated in \cite{Wilt1995, Wilt2003}. For a signal to be detectable, we require that (a) the secondary signal, that is, the difference between the data measured with and without the fractures present, be above the noise floor of the receivers, and (b) that the secondary signal comprises a significant percentage of the primary (model without the fractures). \cite{Marsala2015a} demonstrated noise levels of $0.5 \times 10^-5$ nT and 0.5\% repeatability between data profiles for a cross-well survey conducted between two open-hole horizontal wells, citing that ``this was the lowest noise ever recorded with a crosswell EM system.'' Using this as an upper-bound on our expectations, we adopt a more conservative noise level of $1\times10^-4$ nT and select a percent-threshold of $5\%$. Thus, we consider signal above $1\times10^{-4}$ nT and comprising $>5\%$ of the primary as detectable.

Figure \ref{fig:crosswell_data100} shows the simulated secondary magnetic field data computed for (a) the anisotropic model and (b) the isotropic model at 100 Hz. The top row shows the real component and the bottom row shows the imaginary component. The x-axis shows the transmitter location relative to the center of the fracture, the y-axis shows the receiver location, and the color indicates the secondary magnetic field with respect to a 0.1 S/m whole-space. Regions of the plot which are colored and do not covered by the semi-transparent overlay indicate signal which is both above the noise floor and comprises a significant percentage of the primary.


\input{figures/phys_prop_model/crosswell_data100}

As might be expected, when the transmitter and receiver are centered with the fracture, the secondary signal reaches its maximum amplitude in both the real and imaginary components. The anisotropic model produces stronger signals; the geometry of the survey is perfectly coupled to the radial components of the conductivity ($\sigma_y$ and $\sigma_z$) as the resultant currents and electric fields are purely azimuthal in orientation. An isotropic simulation using $\sigma = 5$ S/m produces identical results to the anisotropic model for this survey geometry. Interestingly, the region over which data is measurable differs between the real and imaginary components. The real component is measurable over a larger range of transmitter positions, while the imaginary component is measurable over a larger range of receiver locations. The partition of EM energy into the real and imaginary components is a function of the conductivity of the medium as well as the frequency. For example, if we increase the frequency to 200 Hz, the amplitude of the imaginary component is larger than the real, and it is measurable over a larger set of source-receiver combinations, as shown in Figure \ref{fig:crosswell_data200}.

\input{figures/phys_prop_model/crosswell_data200}

For the simple setup shown here, there are a significant number of data that can be collected in a cross-well survey which are sensitive to both the anisotropic and isotropic fracture models. In both the 100 Hz and 200 Hz surveys, the extent over which either or both of the real and imaginary components of the data are above the noise floor and comprise a significant percentage of the primary field spanned ~100m for the transmitter location and $> 800$m for the receiver locations. These results providing confidence that EM is a potentially diagnostic tool for characterizing the propped region of a reservoir. Naturally, the success or not of EM in a given setting will depend upon the background conductivity, volume and conductivity of the proppant and fluid, as well as noise conditions in the field. Thus, forward modelling tailored to the setting is required to assess feasibility.
\section{Conclusion}

The large range of scales involved in describing a propped, fractured volume of rock as a geophysical target for an electromagnetic survey poses a challenge for performing numerical simulations. We approached this problem by using effective medium theory to estimate the electrical conductivity of a fracture volume of rock in two steps. First, we estimated the conductivity of a slurry containing proppant and fluid and demonstrated several approaches for creating an electrically conductive slurry, including using conductive particles (or particles coated with a conductive material) as the proppant or including conductive particles, which might be elongated to enhance the conductivity at lower concentrations. General agreement with the values published in the laboratory experiment by \cite{Zhang2016} gives confidence that the estimates produced using effective medium theory are representative of the conductivities that can be achieved; a more rigorous comparison would require control on the conductivity of each of the constituents. In the second step, we estimated the effective conductivity of a fractured volume of rock by assuming that the fracture is composed of a collection of ellipsoidal cracks. These cracks may be preferentially oriented, if simple planar fractures are expected, or randomly oriented if a more complex fracture network is anticipated.

Using this approach, we developed isotropic and anisotropic electrical conductivity models considering proppant and fluid volumes that are representative of current hydraulic fracture operations and simulated a crosswell EM survey. In this example, the fractured volume was detectable under noise estimates that have demonstrated to be achievable in horizontal wells for a waterflood monitoring experiment. This provides evidence that EM has the potential to be a suitable method for characterizing the propped region of a hydraulically fractured reservoir.

The examples examined here neglect the effects of steel-cased well on EM signals. In the vast majority of settings where hydraulic fracturing is conducted, the wells are cased with steel, and therefore, this assumption needs to be addressed. For induction problems, the concern is attenuation of the electromagnetic signals, as eddy currents may be induced in the steel cased well. However, there is increasing interest in using the steel casing as an ``extended electrode'' to help deliver current to depth. In the following chapters, we develop a strategy for including steel cased wells in the EM numerical simulations and examine their influence on EM signals.



